Disclosing funding sources for open access publication fees: the Open APC

### 2016-05-25 article.Rmd
research questions and methods re-tooled, calculations need to be addressed

---
title: 'Disclosing funding sources for open access publication fees: the Open APC initiative'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek (TIB) - German National Library of Science and Technology, Hannover, Germany
date: '7 April 2016'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl
abstract: Publication fees in open access publishing hold a prominent place on the agenda of researchers, policy-makers, and academic publishers. This paper contributes to the evolving empirical basis on open access funding. It describes the Open APC initiative, in which German universities and research organizations share their expenditures for publication fees. As method, the initiative uses existing open data tools to aggregate and disseminate institutional spending on open access publication fees. In total, 29 German research organizations self-reported funding of 6,279 open access journal articles, which amounted to 8,039,339 €. The average payment for each article was 1,280 €, and the median payment 1,209 €. Our data-set comprises only 53 articles in hybrid journals. With an indexing coverage of 99 %, the findings reveal that the DOI agency CrossRef provides both comprehensive bibliographic coverage of the funded open access journal literature and disambiguated names of journal titles and publishing houses. We show that authority control of these bibliographic information is particularly relevant for the comparative study of the economical effects of open access publishing.
---

@@

Drawing on self-reported cost data, we asked to what extend German universities and research institutions covered publication fees for open access journal articles over the last years. More specifically, we addressed three questions that arose while studying fee-based open access publishing:

- how to gather bibliographic metadata about sponsored open access journal articles,
- how is institutional publication fee spending distributed over publishers and journal titles,
- what is the share of sponsored articles in open access journals versus hybrid journals.

@@

The study examined cost information provided by the Open APC initiative. Under this umbrella, German universities and research institutions shared details of how much open access journals charged authors as open data. We analysed the data-set that was released on 13 May 2016 and is stored as version v2.4.2 in the corresponding GitHub repository. At the time of our study, the data-set reported the amount institutions paid for each article including value added tax, and the year an article was invoiced. Bibliographic metadata consisted of the Digital Object Identifier (DOI) to uniquely identify an article, the publisher, and information about the journal titles. The Open APC initiative automatically retrieved this publisher and journal information from CrossRef, a DOI minting agency for scholarly literature. Furthermore, the data-set differentiated whether an open access article was published in a subscription based journal ("hybrid journal").

For gathering normalized publisher and journal titles, we re-aggregated the bibliographic metadata from CrossRef using the DOI on 19 May 2016 because of two reasons: First, CrossRef provides incremental metadata updates to reflect the dynamic landscape of academic publishing including ongoing mergers of publishing houses or name changes. A prominent example in this regard is the merger of the two large publishing houses Springer Business + Media and Nature Publishing that now operate under "Springer Nature". In contrast to this, the Open APC initiative normalized publisher and journal titles based on CrossRef shortly after the institution had provided its cost data. Therefore, re-aggregating the data-set ensures unique reference to currently existing publishers and journals at the expense of information about to which publisher a journal belonged while a sponsored article was published.

Secondly, CrossRef is not the only registration agency for DOI's, but also DataCite and Medra mint DOIs for scholarly work. While obtaining bibliographic information from CrossRef, we also examined the number of publications for which each of the DOI registration agency provided metadata during our study. As a client, we used the R package rcrossref [@rcrossref] developed and maintained by the rOpenSci initiative.[^6] With the function `cr_cn`, we requested the linking resource `application/vnd.crossref.unixsd+xml`, which main function is to support text mining activities.[^5] The advantage of this XML format was that it distinguishes full and abbreviated journal titles as well as the media types of ISSN's, the International Standard Serial Number used to identify journals. It also contained license information and disambiguated publisher information, thus avoiding confusion about licensing and naming of publisher houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values.

After summarizing publication fee spending per institution, publisher and journal, we compared the share of hybrid journal articles in the Open APC data-set with those of Wellcome Trust, FWF and Jisc. For this aim, we systematically reviewed the proportions for each year from the accompanying literature.

@@

```{r, echo=FALSE, cache = FALSE}
# load data
my_apc <- read.csv("data/apc_de.csv", header = TRUE, sep =",", stringsAsFactors = FALSE)

# data cleaning
library(dplyr)
# remove Bielefelds Zero value
my_apc <- filter(my_apc, euro > 0)
# remove intech book chapters
my_apc <-  filter(my_apc, !is.na(journal_full_title))
# clean publisher titles, because metadata could not be fetched
my_apc$publisher[grep("BioMed Central", my_apc$publisher, ignore.case = T)] <- "Springer Science + Business Media"
my_apc$publisher[grep("Impact Journals, LLC", my_apc$publisher, ignore.case = T)] <- "Impact Journals LLC"
my_apc$publisher[grep("e-Century Publishing Corporation", my_apc$publisher, ignore.case = T)] <- "e-Century Publ. Corp."
my_apc$publisher[grep("Academic Journals", my_apc$publisher, ignore.case = T)] <- "Academic Journals"
# fix plos name change in journal titles
my_apc$journal_full_title <- gsub("PLoS", "PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PloS","PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PlOS","PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Sustainibility","Sustainability", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Surgeal","Surgical", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Oncotarget","OncoTarget", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("BMC cancer","BMC Cancer", my_apc$journal_full_title)

@@

On February 19th 2016,[^16a] the Open APC initiative covered `r nrow(my_apc)` articles, whose publication fees were centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions. The number of supported open access journal articles, which were reported to this initiative, grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. Because of a time lag between payments made and reporting them to the Open APC initiative, only `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institution were able to partly contribute their cost data for 2015 at the time of this analysis.

```{r articles_year, fig.width=6, fig.height=2.5,fig.cap="Growth of Open APC Initiative"}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())
ggplot(apc_time, aes(factor(period), cumsum(articles), group = 1)) +
  geom_line(stat = "identity", colour = "blue") +
  geom_point() +
  xlab("Year") +
  ylab("OA articles reported\n(cumulative sum)") +
  scale_y_continuous(limits = c(0,8000))
```

@@

```{r payments_year, fig.width=6, fig.height=3,fig.cap="Payments per year"}
ggplot(my_apc, aes(factor(period), euro)) +
  geom_boxplot() +
  xlab("Year") +
  ylab("Publishing fee in EURO(log scale)") +
  scale_y_continuous(limits = c(0,8000))
```

@@

Along with the price information, participating institutions were required to identify funded articles by their DOI. They were reported for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. Of those, `r my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow()` were indexed in CrossRef, representing `r my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow() / my_apc %>% nrow() * 100`  % of all funded publications. The reasons why articles identified by a DOI were not registered with CrossRef differed. Some journals were not indexed by CrossRef at the time of our study but by the DOI agencies DataCite (Journal of new frontiers in spatial concepts published by KIT Scientific Publishing) and Medra (DIE ERDE: Journal of the Geographical Society of Berlin). In other cases, either the DOI did not refer to the full text despite the fact that the journal was indexed on a regular basis (compare <http://doi.org/10.1186/1471-2105-13-S19-S7> with <http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-13-S19-S7>) or the resource type could not be retrieved, although the DOI resolves (<http://doi.org/10.1186/s12885-015-1795-7>).

@@

```{r}
apc_publisher <- my_apc
apc_publisher$publisher <- factor (apc_publisher$publisher, levels = c(rownames(data.frame(rev(sort(table(apc_publisher$publisher)))))))

levels(apc_publisher$publisher)[11:length(levels(apc_publisher$publisher))] <- "other"
```

We used the DOI to automatically fetch publisher and journal names for each article from the CrossRef REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to  `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

```{r results='asis'}
tt <- apc_publisher %>%
  group_by(publisher) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro))

colnames(tt) <- c("Publisher", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')
```

@@

```{r results='asis'}
apc_journal_full_title <- my_apc
apc_journal_full_title$journal_full_title <- factor (apc_journal_full_title$journal_full_title, levels = c(rownames(data.frame(rev(sort(table(apc_journal_full_title$journal_full_title)))))))

levels(apc_journal_full_title$journal_full_title)[11:length(levels(apc_journal_full_title$journal_full_title))] <- "other"

tt <- apc_journal_full_title %>%
  group_by(journal_full_title) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro))

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per journal (in €)', justify = 'lrrrrrr')
```

Prices also varied within single journals. Based on the number of articles paid for, Table 5 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these ten journals represent `r apc_journal_full_title %>% filter(!journal_full_title == "other") %>% nrow() / my_apc %>% nrow() * 100` % of all payments. In the case of Atmospheric Chemistry and Physics Discussions, the price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

>>>>

---
title: 'Disclosing funding sources for open access publication fees: the Open APC
  initiative'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University (TIB) - German National
  Library of Science and Technology, Hannover, Germany
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek
date: "7 April 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl
abstract: Publication fees in open access publishing hold a prominent place on the
  agenda of researchers, policy-makers, and academic publishers. This paper contributes
  to the evolving empirical basis on open access funding. It describes the Open APC
  initiative, in which German universities and research organizations share their
  expenditures for publication fees. As method, the initiative uses existing open
  data tools to aggregate and disseminate institutional spending on open access publication
  fees. In total, 29 German research organizations self-reported funding of 6,279
  open access journal articles, which amounted to 8,039,339 €. The average payment
  for each article was 1,280 €, and the median payment 1,209 €. Our data-set comprises
  only 53 articles in hybrid journals. With an indexing coverage of 99 %, the findings
  reveal that the DOI agency CrossRef provides both comprehensive bibliographic coverage
  of the funded open access journal literature and disambiguated names of journal
  titles and publishing houses. We show that authority control of these bibliographic
  information is particularly relevant for the comparative study of the economical
  effects of open access publishing.
---

@@

The aim of this study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Drawing on self-reported data that German institutions openly shared within the Open APC initiative, we focused our analysis on the amount of institutional publication fee spending, and compared these findings with that of related efforts in Austria and the UK. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We used cost data released by the Open APC initiative on May 13, 2016, to analyse expenditures on publication fees by German universities and research organisations. In addition to administrative data about the amount paid per article including value added tax and the year of invoicing, the Open APC initiative has begun to share information about whether an article was published in a fully or hybrid open access journal. Bibliographic metadata recorded in the data-set consisted of the Digital Object Identifier (DOI) to uniquely identify an article, and information about publisher and journal titles that were automatically retrieved from CrossRef when available.

Data collection in this study involved obtaining cost data from related open data initiatives and retrieving bibliographic metadata updates. To compare self-reported spending on open access publishing by Germany universities and research organisations with that of other initiatives, we reviewed  the openly available spreadsheets provided by the Wellcome Trust [@wellcome_apc], the Austrian Science Fund FWF [@fwf_apc] and Jisc. All data-sets were openly available through a figshare collection. We obtained the size  of funding for all publications and fully open access journal articles. The Euro foreign exchange reference rates from the European Central bank were consulted to convert costs reported in GBP to Euro.

To provide a current view on open access publishing, DOIs reported in the Open APC data-set served us as a basis to re-collect bibliographic metadata from CrossRef on May 19, 2016. Since the beginning of data normalisation using CrossRef by the Open APC initiative in 2014, CrossRef has provided incremental metadata updates in which ongoing mergers of publishing houses or name changes are reflected. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing announced on May 6, 2015, that now operate under Springer Nature. In contrast to this, the Open APC initiative gathered publisher and journal titles shortly after an institution provided its expenditures.

As a client to access the CrossRef web service,  we used the R package rcrossref [@rcrossref] developed and maintained by the rOpenSci initiative.[^6] With the function `cr_cn`, we requested the linking resource `application/vnd.crossref.unixsd+xml` used  to support text mining activities.[^5] At the time of our study, this XML format distinguished full and abbreviated journal titles as well as the media types of ISSN's, the International Standard Serial Number used to identify journals. It also contained disambiguated publisher information, thus avoiding confusion about naming of publisher houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Since CrossRef is not the only registration agency for DOI's, but also the agencies DataCite and Medra mint DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

- statistical calculations

@@

```{r, echo=FALSE, cache = FALSE}
# load data
my_apc <- jsonlite::stream_in(file("data/cr-apc-all.json"), verbose = FALSE)

# data cleaning
library(dplyr)
## remove Bielefelds Zero value
my_apc <- dplyr::filter(my_apc, euro > 0)
# remove intech book chapters
my_apc <-  dplyr::filter(my_apc, !is.na(journal_full_title))

# fix plos name change in journal titles
my_apc$journal_full_title <- gsub("PLoS", "PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PLOS ONE", "PLOS ONE", my_apc$journal_full_title)

# remove Austrian FWF data
my_apc <- dplyr::filter(my_apc, !institution == "FWF - Austrian Science Fund")
# remove data from 2016
my_apc <- dplyr::filter(my_apc, !period == "2016")

@@

On May 13, 2016,[^16a] the Open APC initiative covered `r nrow(my_apc)` articles whose publication fees were centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions until 2015. The number of supported open access journal articles grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions contributed their cost data for 2015 at the time of this analysis, suggesting that there exits a time lag between payments made and reporting these spending to the Open APC initiative.

```{r histm, fig.width=6, fig.height=2.5,fig.cap="Histogram APC"}
require(ggplot2)
## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))

ggplot(my_apc, aes(euro)) + 
  geom_histogram(binwidth = 250, colour ="black", boundary = 0) + 
  scale_x_continuous(limits = c(0,8000)) +
  xlab("Publication fee in €") +
  ylab("Articles") +
  theme_bw() + science_theme

```{r articles_year, fig.width=6, fig.height=2.5,fig.cap="Growth of Open APC Initiative"}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())

## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))

ggplot(apc_time, aes(factor(period), cumsum(articles), group = 1)) +
  geom_line(stat = "identity", colour = "blue") +
  geom_point() +
  xlab("Year") +
  ylab("OA articles reported\n(cumulative sum)") +
  scale_y_continuous(limits = c(0,8000)) +
  theme_bw() + science_theme
ggsave("figure_1.pdf", height=4, width=6)

@@

```{r payments_year, fig.width=6, fig.height=3,fig.cap="Payments per year"}
ggplot(my_apc, aes(factor(period), euro)) +
  geom_boxplot(outlier.size = 0.5) +
  xlab("Year") +
  ylab("Publication fee paid in EURO") +
  scale_y_continuous(limits = c(0,8000)) + 
  theme_bw() + science_theme
ggsave("figure_2.pdf", height=4, width=6)

@@

Along with the price information, participating institutions were required to identify funded articles by their DOI. They were reported for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. Of those, `my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow()` were indexed in CrossRef, representing `my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow() / my_apc %>% nrow() * 100`  % of all funded publications. The reasons why articles identified by a DOI were not registered with CrossRef differed. Some journals were not indexed by CrossRef at the time of our study but by the DOI agencies DataCite (Journal of new frontiers in spatial concepts published by KIT Scientific Publishing) and Medra (DIE ERDE: Journal of the Geographical Society of Berlin). In other cases, either the DOI did not refer to the full text despite the fact that the journal was indexed on a regular basis (compare <http://doi.org/10.1186/1471-2105-13-S19-S7> with <http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-13-S19-S7>) or the resource type could not be retrieved, although the DOI resolves (<http://doi.org/10.1186/s12885-015-1795-7>).

@@

```{r}
apc_publisher <- my_apc
publisher_ <- names(rev(sort(table(apc_publisher$publisher)))[1:10])
apc_publisher$publisher <- ifelse(apc_publisher$publisher %in% publisher_, apc_publisher$publisher, "Other")
# SORT FACTOR
apc_publisher$publisher <- factor(apc_publisher$publisher, levels=c(publisher_, "Other")) 
```

We used the DOI to automatically fetch publisher and journal names for each article from the CrossRef REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to  `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

```{r results='asis'}
tt <- apc_publisher %>%
  group_by(publisher) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), SD = sd(euro), Median = median(euro), Min = min(euro), Max = max(euro)) %>%
  mutate(range_apc = paste(Min, "--", Max)) %>%
  select(-Min, -Max)

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')
```

@@

```{r results='asis'}
apc_journal <- my_apc

journal_ <- names(rev(sort(table(apc_journal$journal_full_title)))[1:10])
apc_journal$journal_full_title <- ifelse(apc_journal$journal_full_title %in% journal_, 
                                         apc_journal$journal_full_title, "Other")

# SORT FACTOR
apc_journal$journal_full_title <- factor(apc_journal$journal_full_title, levels=c(journal_, "Other")) 
tt <- apc_journal %>%
  group_by(journal_full_title) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), SD = sd(euro), Median = median(euro), Min = min(euro), Max = max(euro)) %>%
  mutate(range_apc = paste(Min, "--", Max)) %>%
  select(-Min, -Max)

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per journal (in €)', justify = 'lrrrrrr')
```

Prices also varied within single journals. Based on the number of articles paid for, Table 5 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these ten journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of all payments. In the case of Atmospheric Chemistry and Physics Discussions, the price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.
